---
title: "**Guild Assignment With INVGuild**"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## **Load Required Libraries**

```{r}
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(ggpubr)
```

## **Read Input Files**

Two input files are required:

* the file containing the INVGuild database (*INVGuild.tsv*)
* the file containing the list of invertebrate taxa (formatted like [SILVA](https://www.arb-silva.de/)) to be assigned to a trophic guild in a column named "Taxon"

```{r}
db <- read.csv("INVGuild.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
tax <- read.csv("Taxonomy_LUCAS_Metazoa.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

## **Helper Function and Variables**

```{r}
extract_tax_level <- function(df, level_prefix) {
  pattern <- paste0(level_prefix, "__[^;]+")
  df %>%
    mutate(last_part = str_extract(Taxon, pattern))
}

levels <- c("Species", "Genus", "Family", "Order", "Class")
lvl_prefix <- c("s", "g", "f", "o", "c")
```

## **Clean and Prepare the Database**

```{r}
# Remove entries with missing ecological or taxonomic level data
db_clean <- db[!is.na(db$Ecology) & !is.na(db$Taxonomic.level) & db$Ecology != "", ]
# Split the database into subsets based on the taxonomic level of guild assignment
db_list <- setNames(lapply(levels, function(lvl) {db_clean[db_clean$Taxonomic.level == lvl, ]}), levels)
# Replace white spaces with underscores in species-level taxon names
db_list$Species$Taxon <- str_replace_all(db_list$Species$Taxon, "\\s+", "_")
# Extract species level IDs
db_list$Species <- db_list$Species %>%
  mutate(last_part = str_extract(Taxon, "s__.*"))
# Extract genus, family, order, and class level IDs
for (i in 2:length(db_list)) {
  db_list[[i]] <- extract_tax_level(db_list[[i]], lvl_prefix[i])
}
```

## **Assign Ecological Guilds**

```{r}
Assigned <- list()

# Assignment at the species level
TaxToBeAssigned <- tax %>%
  mutate(last_part = str_extract(Taxon, "s__.*"))
res <- TaxToBeAssigned %>%
  left_join(db_list$Species %>% select(last_part, Ecology), by = "last_part")
Assigned$Species <- res %>% filter(!is.na(Ecology))
TaxToBeAssigned <- res %>% filter(is.na(Ecology))

# Assignment at the genus level
TaxToBeAssigned <- extract_tax_level(TaxToBeAssigned[, 1:4], "g")
res <- TaxToBeAssigned %>%
  left_join(db_list$Genus %>% select(last_part, Ecology), by = "last_part")
Assigned$Genus <- res %>% filter(!is.na(Ecology))
TaxToBeAssigned <- res %>% filter(is.na(Ecology))

# Assignment at the family level
TaxToBeAssigned <- extract_tax_level(TaxToBeAssigned[, 1:4], "f")
res <- TaxToBeAssigned %>%
  left_join(db_list$Family %>% select(last_part, Ecology), by = "last_part")
Assigned$Family <- res %>% filter(!is.na(Ecology))
TaxToBeAssigned <- res %>% filter(is.na(Ecology))

# Assignment at the order level
TaxToBeAssigned <- extract_tax_level(TaxToBeAssigned[, 1:4], "o")
res <- TaxToBeAssigned %>%
  left_join(db_list$Order %>% select(last_part, Ecology), by = "last_part")
Assigned$Order <- res %>% filter(!is.na(Ecology))
TaxToBeAssigned <- res %>% filter(is.na(Ecology))

# Assignment at the class level
TaxToBeAssigned <- extract_tax_level(TaxToBeAssigned[, 1:4], "c")
res <- TaxToBeAssigned %>%
  left_join(db_list$Class %>% select(last_part, Ecology), by = "last_part")
Assigned$Class <- res %>% filter(!is.na(Ecology))
TaxToBeAssigned <- res %>% filter(is.na(Ecology))
```

## **Clean and Export Final Results**

```{r}
# Clean the output and add assignment level
for (i in 1:length(Assigned)) {
  Assigned[[i]] <- Assigned[[i]] %>%
    select(-last_part) %>%
    mutate(AssignmentLevel = levels[i])
}
Assigned_df <- do.call(rbind, Assigned)

# Export results
write.table(Assigned_df, "INVGuild_assigned.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(TaxToBeAssigned[1:3], "INVGuild_unassigned.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

# Check how many entries received a trophic guild assignment
paste0(round(100*(nrow(Assigned_df) / (nrow(Assigned_df)+nrow(TaxToBeAssigned))),2),
       "% of the original entries were assigned to a trophic guild")
```

## **Visualize The Classification Results**

```{r}
# Prepare data for visualization
SumTab <- as.data.frame.matrix(table(Assigned_df$Ecology, Assigned_df$AssignmentLevel))
desired_order <- c("Class", "Order", "Family", "Genus", "Species")
SumTab <- SumTab[, desired_order[desired_order %in% colnames(SumTab)]]
SumTab$Ecology <- row.names(SumTab)
SumTab_long <- SumTab %>%
  pivot_longer(cols = -Ecology, names_to = "AssignmentLevel", values_to = "Count")
SumTab_long$AssignmentLevel <- factor(SumTab_long$AssignmentLevel,
                                      levels = rev(c("Class", "Order", "Family", "Genus", "Species")))
SumTab_long$All <- "All levels"

# Get the number of entries assigned per taxonomic level
lbls <- c(
  paste0("species\n\n(n=", sum(Assigned_df$AssignmentLevel == "Species"), ")"),
  paste0("genus\n\n(n=", sum(Assigned_df$AssignmentLevel == "Genus"), ")"),
  paste0("family\n\n(n=", sum(Assigned_df$AssignmentLevel == "Family"), ")"),
  paste0("order\n\n(n=", sum(Assigned_df$AssignmentLevel == "Order"), ")"),
  paste0("class\n\n(n=", sum(Assigned_df$AssignmentLevel == "Class"), ")")
)

# Barplots showing the relative abundance of trophic guilds identified in the full dataset (on the left) and per taxonomic level (on the)
ggarrange(
  ggplot(SumTab_long, aes(x = All, y = Count, fill = Ecology)) +
    geom_bar(position="fill", stat = "identity") +
    labs(x = "", y = "") +
    theme_minimal() +
    labs(fill = NULL),
  ggplot(SumTab_long, aes(x = AssignmentLevel, y = Count, fill = Ecology)) +
    geom_bar(position="fill", stat = "identity") +
    labs(x = "", y = "") +
    scale_x_discrete(labels = lbls) +
    theme_minimal() +
    labs(fill = NULL),
  ncol=2, widths = c(1,2), common.legend = TRUE
)
```
